### 2020-10-15 priority-cp.tex
Compressed lists and subsubsections, minor rewrites in text.

From the perspective of process calculus, commuting conversions behave strangely. Let us consider the commuting conversion for $\pcp{\tm{\recv{x}{y}{P}}}$:
\[
  \usingnamespace{pcp}
  (\kappa_{\parr})
  \quad
  \tm{\res{z}{z'}{(\ppar{\recv{x}{y}{P}}{Q})}}
  \red
  \tm{\recv{x}{y}{\res{z}{z'}{(\ppar{P}{Q})}}}
\]
As a result of $(\kappa_{\parr})$, $\pcp{\tm{Q}}$ becomes blocked on $\pcp{\tm{\labrecv{x}{y}}}$, and any actions $\pcp{\tm{Q}}$ was able to perform become unavailable. Consequently, CP is non-confluent, for instance:
\begin{mathpar}
  \usingnamespace{pcp}
  \setlength{\arraycolsep}{2em}
  \begin{array}{cc}
    \multicolumn{2}{c}{%
    \hspace*{10ex}
    {\tm{\res{x}{x'}{(\ppar{\recv{a}{y}{P}}{\res{z}{z'}{(\ppar{\close{z}{\halt}}{\wait{z'}{Q}})}})}}}}
    \\
    \qquad\rotatebox[origin=c]{215}{$\red$}
    &
    \rotatebox[origin=c]{325}{$\red$}\qquad
    \\
    {\tm{\recv{a}{y}{\res{x}{x'}{(\ppar{P}{\res{z}{z'}{(\ppar{\close{z}{\halt}}{\wait{z'}{Q}})}})}}}}
    &
    {\tm{\res{x}{x'}{(\ppar{\recv{a}{y}{P}}{Q})}}}
    \\ 
    &
    \rotatebox[origin=c]{270}{$\red$}
    \\
    &
    \tm{\recv{a}{y}{\res{x}{x'}{(\ppar{P}{Q})}}}
  \end{array}
\end{mathpar}
Commuting conversions break our intuition for priorities, since ``if a prefix on a channel endpoint $\pcp{\tm{x}}$ with priority $\cs{o}$ is pulled out at top level, then to preserve priority conditions in the typing rules [..], it is necessary to increase priorities of all actions after the prefix on $\pcp{\tm{x}}$'' by $\cs{o+1}$~\cite{dardhagay18}. Consequently, the intuition that an action with lower priority occurs before an action with higher priority is shattered.

>>>>

From the perspective of process calculus, commuting conversions behave strangely. Let us consider the commuting conversion for $\pcp{\tm{\recv{x}{y}{P}}}$:
\[
  \usingnamespace{pcp}
  (\kappa_{\parr})
  \quad
  \tm{\res{z}{z'}{(\ppar{\recv{x}{y}{P}}{Q})}}
  \red
  \tm{\recv{x}{y}{\res{z}{z'}{(\ppar{P}{Q})}}}
\]
As a result of $(\kappa_{\parr})$, $\pcp{\tm{Q}}$ becomes blocked on $\pcp{\tm{\labrecv{x}{y}}}$, and any actions $\pcp{\tm{Q}}$ was able to perform become unavailable. Consequently, CP is non-confluent. For instance,
\begin{mathpar}
  \usingnamespace{pcp}
  \setlength{\arraycolsep}{2em}
  \begin{array}{cc}
    \multicolumn{2}{c}{%
    \hspace*{10ex}
    {\tm{\res{x}{x'}{(\ppar{\recv{a}{y}{P}}{\res{z}{z'}{(\ppar{\close{z}{\halt}}{\wait{z'}{Q}})}})}}}}
    \\
    \qquad\rotatebox[origin=c]{215}{$\red$}
    &
    \rotatebox[origin=c]{325}{$\red^2$}\qquad
    \\
    {\tm{\recv{a}{y}{\res{x}{x'}{(\ppar{P}{\res{z}{z'}{(\ppar{\close{z}{\halt}}{\wait{z'}{Q}})}})}}}}
    &
    {\tm{\recv{a}{y}{\res{x}{x'}{(\ppar{P}{Q})}}}}
  \end{array}
\end{mathpar}
Commuting conversions break our intuition that an action with lower priority occurs before an action with higher priority, since ``if a prefix on a channel endpoint $\pcp{\tm{x}}$ with priority $\cs{o}$ is pulled out at top level, then to preserve priority conditions in the typing rules [..], it is necessary to increase priorities of all actions after the prefix on $\pcp{\tm{x}}$'' by $\cs{o+1}$~\cite{dardhagay18}.

### 2020-10-15 priority-gv.tex
Compressed lists and subsubsections, minor rewrites in text.

\subsubsection*{Operational Semantics}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
\begin{array}[t]{lcl}
    \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  & \Coloneqq & \tm{\phi\;M}
    \sep        \tm{\ppar{\conf{C}}{\conf{D}}}
    \sep        \tm{\res{x}{x'}{\conf{C}}}
  \\
  \tm{\phi}
  & \Coloneqq & \tm{\main}
    \sep        \tm{\child}
\end{array}
\]
Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
\begin{array}{ll}
  \tm{\main}  + \tm{\child} = \tm{\main}
  \\
  \tm{\child} + \tm{\main}  = \tm{\main}
\end{array}
\quad
\begin{array}{ll}
  \tm{\child} + \tm{\child} = \tm{\child}
  \\
  \tm{\main}  + \tm{\main} \; \text{undefined}
\end{array}
\]
Values ($\tm{V}$, $\tm{W}$), evaluation contexts ($\tm{E}$), thread evaluation contexts ($\tm{\conf{F}}$), and configuration contexts ($\tm{\conf{G}}$) are defined by the following grammars:
\[
\begin{array}[t]{lcl}
  \tm{V}, \tm{W}
  & \Coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{\unit}
    \sep        \tm{\pair{V}{W}}
    \sep        \tm{\inl{V}}
    \sep        \tm{\inr{V}}
  \\
  \tm{E}
  & \Coloneqq & \tm{\hole} \\
  & \sep      & \tm{E\;M}
    \sep        \tm{V\;E} \\
  & \sep      & \tm{\letunit{E}{N}} \\
  & \sep      & \tm{\pair{E}{M}}
    \sep        \tm{\pair{V}{E}}
    \sep        \tm{\letpair{x}{y}{E}{M}} \\
  & \sep      & \tm{\inl{E}}
    \sep        \tm{\inr{E}}
    \sep        \tm{\casesum{E}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{E}}
  \\
  \tm{\conf{F}}
  & \Coloneqq & \tm{\phi\;E}
  \\
  \tm{\conf{G}}
  & \Coloneqq & \tm{\hole}
    \sep        \tm{\ppar{\conf{G}}{\conf{C}}}
    \sep        \tm{\res{x}{y}{\conf{G}}}
\end{array}
\]
We factor the reduction relation of PGV into a deterministic reduction on terms ($\tred$) and a non-deterministic reduction on configurations ($\cred$), see \cref{fig:pgv-operational-semantics}. We write $\tred^+$ and $\cred^+$ for the transitive closures, and $\tred^\star$ and $\cred^\star$ for the reflexive-transitive closures.

@@

While our configuration reduction is based on the standard evaluation for GV, the increased expressiveness of PGV allows us to simplify the relation on two counts:

Firstly, we decompose the $\tm{\fork}$ construct. In GV, $\tm{\fork}$ creates a new channel, spawns a child thread, and, when the child thread finishes, it closes the channel to its parent. In PGV, these are three separate operations: $\tm{\new}$, $\tm{\spawn}$, and $\tm{\close}$. We no longer require that every child thread finishes by returning a terminated channel. Consequently, we also simplify the evaluation of the $\tm{\link}$ construct. In GV, $\tm{\link}$ needs to return a terminated channel after evaluating the link. In PGV, the behaviour matches that of link in CP.

Secondly, our structural congruence is type preserving. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well-typed configuration using $\equiv$.

@@

While our configuration typing is based on the standard typing for GV, it differs on two counts:

Firstly, we require that child threads return the unit value, as opposed to a terminated channel.

Secondly, we simplify typing for parallel composition. In GV, each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$ across the parallel composition. Consequently, associativity of parallel composition does not preserve typing. Unfortunately, splitting exactly one channel across a parallel composition in GV plays a crucial role in guaranteeing deadlock freedom. In PGV, we guarantee deadlock freedom using priorities, which removes the channel pseudo-type $\ty{S^\sharp}$, simplifies typing for parallel composition, and restores type preservation for the structural congruence.

@@

\subsubsection*{Progress and deadlock freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well-typed PGV term does not necessarily produce \emph{just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV. Let us consider an example, adapted from Lindley and Morris~\cite{lindleymorris15}, in which a term returns an endpoint linked to an echo server. The echo server receives a value and sends it back unchanged:
\[
  \tm{\echo_x} \defeq \tm{\letpair{y}{x}{\recv\;{x}}{\letbind{x}{\send\;\pair{y}{x}}{\close\;x}}}
\]
Consider the program which creates a new channel, with endpoints $\tm{x}$ and $\tm{x'}$, spawns off an echo server listening on $\tm{x'}$, and then returns $\tm{x}$:
\[
  \tm{\main\;\letpair{x}{x'}{\new}{\andthen{\spawn\;(\lambda\unit.\echo_{x'})}}{x}}
\]
If we reduce the above program, we get the following normal form. Clearly, no more evaluation is possible, even though the configuration contains the thread $\tm{\child\;\echo_{x'}}$, which is blocked on $\tm{x'}$:
\[
  \tm{\res{x}{x'}{(\ppar{\main\;x}{\child\;\echo_{x'}})}}
\]
In~\cref{cor:pgv-closed-progress} we show that if a term does not return an endpoint, it must produce \emph{only} a value.

@@

Canonical forms deviate from those for GV, in that we opt to move all $\nu$-binders to the top. The standard GV normal forms, alternating $\nu$-binders and their corresponding parallel compositions, do not work for PGV, since multiple channels may be split across a single parallel composition:
\begin{definition}[Canonical Forms]
  \label{def:pgv-canonical-forms}
  A~configuration $\tm{\conf{C}}$ is in canonical form if it is of the following form, where no term $\tm{M_i}$ is a value:
  \[
    \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;N)}}}
  \]
\end{definition}
\begin{restatablelemma}{lempgvcanonicalforms}[Canonical Forms]
  \label{lem:pgv-canonical-forms}
  If $\cseq[\main]{\ty{\Gamma}}{\conf{C}}$, there exists some $\tm{\conf{D}}$ such that $\tm{\conf{C}}\equiv\tm{\conf{D}}$ and $\tm{\conf{D}}$ is in canonical form.
\end{restatablelemma}
\begin{proof}
  We move any $\nu$-binders to the top using \LabTirName{SC-ResExt}, discard any superfluous occurrences of $\tm{\child\;\unit}$ using \LabTirName{SC-ParNil}, and move the main thread to the rightmost position using \LabTirName{SC-ParComm} and \LabTirName{SC-ParAssoc}.
\end{proof}

% For simplicity, we only present progress for closed configurations. A~suitable notion of progress for open configurations does exist, but its normal forms are significantly more complex, as they may also contain threads which are blocked on free endpoints, threads blocked on endpoints for which the duals are still to be used in those threads blocked on free endpoints, \etc.

The statement of closed progress is rather subtle. A~configuration either reduces, or it is equivalent to configuration in normal form. Crucial to the normal form is that each term $\tm{M_i}$ is blocked on the corresponding channel $\tm{x_i}$, and hence no two terms act on dual endpoints. Furthermore, no term $\tm{M_i}$ can perform a communication action by itself, since those are excluded by the definition of actions:
\begin{definition}[Normal Forms]
  A~configuration $\tm{\conf{C}}$ is in normal form if it is of the following form, where each $\tm{M_i}$ is ready to act on $\tm{x_i}$:
  \[
    \tm{\conf{C}}
    \equiv
    \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;V)}}},
  \]
\end{definition}
\begin{restatabletheorem}{thmpgvclosedprogressconfs}[Closed Progress, $\cred$]
  \label{lem:pgv-closed-progress-confs}
  If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}}$ is equivalent to a configuration in normal form.
\end{restatabletheorem}
\input{prf/thm-pgv-closed-progress-confs}

>>>>

\subsubsection*{Operational Semantics}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
\begin{array}[t]{lcl}
    \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  & \Coloneqq & \tm{\phi\;M}
    \sep        \tm{\ppar{\conf{C}}{\conf{D}}}
    \sep        \tm{\res{x}{x'}{\conf{C}}}
  \\
  \tm{\phi}
  & \Coloneqq & \tm{\main}
    \sep        \tm{\child}
\end{array}
\]
Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
  \tm{\main}  + \tm{\child} = \tm{\main}
  \quad
  \tm{\child} + \tm{\main}  = \tm{\main}
  \quad
  \tm{\child} + \tm{\child} = \tm{\child}
  \quad
  (\tm{\main}  + \tm{\main} \; \text{is undefined})
\]
Values ($\tm{V}$, $\tm{W}$), evaluation contexts ($\tm{E}$), thread evaluation contexts ($\tm{\conf{F}}$), and configuration contexts ($\tm{\conf{G}}$) are defined by the following grammars:
\[
\begin{array}[t]{lcl}
  \tm{V}, \tm{W}
  & \Coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{\unit}
    \sep        \tm{\pair{V}{W}}
    \sep        \tm{\inl{V}}
    \sep        \tm{\inr{V}}
  \\
  \tm{E}
  & \Coloneqq & \tm{\hole} \\
  & \sep      & \tm{E\;M}
    \sep        \tm{V\;E} \\
  & \sep      & \tm{\letunit{E}{N}} \\
  & \sep      & \tm{\pair{E}{M}}
    \sep        \tm{\pair{V}{E}}
    \sep        \tm{\letpair{x}{y}{E}{M}} \\
  & \sep      & \tm{\inl{E}}
    \sep        \tm{\inr{E}}
    \sep        \tm{\casesum{E}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{E}}
  \\
  \tm{\conf{F}}
  & \Coloneqq & \tm{\phi\;E}
  \\
  \tm{\conf{G}}
  & \Coloneqq & \tm{\hole}
    \sep        \tm{\ppar{\conf{G}}{\conf{C}}}
    \sep        \tm{\res{x}{y}{\conf{G}}}
\end{array}
\]
We factor the reduction relation of PGV into a deterministic reduction on terms ($\tred$) and a non-deterministic reduction on configurations ($\cred$), see \cref{fig:pgv-operational-semantics}. We write $\tred^+$ and $\cred^+$ for the transitive closures, and $\tred^\star$ and $\cred^\star$ for the reflexive-transitive closures.

@@

While our configuration reduction is based on the standard evaluation for GV, the increased expressiveness of PGV allows us to simplify the relation on two counts.
Firstly, we decompose the $\tm{\fork}$ construct. In GV, $\tm{\fork}$ creates a new channel, spawns a child thread, and, when the child thread finishes, it closes the channel to its parent. In PGV, these are three separate operations: $\tm{\new}$, $\tm{\spawn}$, and $\tm{\close}$. We no longer require that every child thread finishes by returning a terminated channel. Consequently, we also simplify the evaluation of the $\tm{\link}$ construct. In GV, $\tm{\link}$ needs to return a terminated channel after evaluating the link. In PGV, the behaviour matches that of link in CP.
Secondly, our structural congruence is type preserving. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well-typed configuration using $\equiv$.

@@

While our configuration typing is based on the standard typing for GV, it differs on two counts.
Firstly, we require that child threads return the unit value, as opposed to a terminated channel.
Secondly, we simplify typing for parallel composition. In GV, each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$ across the parallel composition. Consequently, associativity of parallel composition does not preserve typing. Unfortunately, splitting exactly one channel across a parallel composition in GV plays a crucial role in guaranteeing deadlock freedom. In PGV, we guarantee deadlock freedom using priorities, which removes the channel pseudo-type $\ty{S^\sharp}$, simplifies typing for parallel composition, and restores type preservation for the structural congruence.

@@

\subsubsection*{Progress and deadlock freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well-typed PGV term does not necessarily produce \emph{just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV. Let us consider an example, adapted from Lindley and Morris~\cite{lindleymorris15}, in which a term returns an endpoint linked to an echo server. The echo server receives a value and sends it back unchanged:
\[
  \tm{\echo_x} \defeq \tm{\letpair{y}{x}{\recv\;{x}}{\letbind{x}{\send\;\pair{y}{x}}{\close\;x}}}
\]
Consider the program which creates a new channel, with endpoints $\tm{x}$ and $\tm{x'}$, spawns off an echo server listening on $\tm{x'}$, and then returns $\tm{x}$:
\[
  \tm{\main\;\letpair{x}{x'}{\new}{\andthen{\spawn\;(\lambda\unit.\echo_{x'})}}{x}}
\]
If we reduce the above program, we get $\tm{\res{x}{x'}{(\ppar{\main\;x}{\child\;\echo_{x'}})}}$. Clearly, no more evaluation is possible, even though the configuration contains the thread $\tm{\child\;\echo_{x'}}$, which is blocked on $\tm{x'}$. In~\cref{cor:pgv-closed-progress} we show that if a term does not return an endpoint, it must produce \emph{only} a value.

@@

Canonical forms deviate from those for GV, in that we opt to move all $\nu$-binders to the top. The standard GV canonical form, alternating $\nu$-binders and their corresponding parallel compositions, does not work for PGV, since multiple channels may be split across a single parallel composition:
\begin{definition}[Canonical Forms]
  \label{def:pgv-canonical-forms}
  A~configuration $\tm{\conf{C}}$ is in canonical form if it is of the following form, where no term $\tm{M_i}$ is a value:
  \[
    \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;N)}}}
  \]
\end{definition}
\begin{restatablelemma}{lempgvcanonicalforms}[Canonical Forms]
  \label{lem:pgv-canonical-forms}
  If $\cseq[\main]{\ty{\Gamma}}{\conf{C}}$, there exists some $\tm{\conf{D}}$ such that $\tm{\conf{C}}\equiv\tm{\conf{D}}$ and $\tm{\conf{D}}$ is in canonical form.
\end{restatablelemma}
\begin{proof}
  We move any $\nu$-binders to the top using \LabTirName{SC-ResExt}, discard any superfluous occurrences of $\tm{\child\;\unit}$ using \LabTirName{SC-ParNil}, and move the main thread to the rightmost position using \LabTirName{SC-ParComm} and \LabTirName{SC-ParAssoc}.
\end{proof}

A~configuration either reduces, or it is equivalent to configuration in normal form. Crucial to the normal form is that each term $\tm{M_i}$ is blocked on the corresponding channel $\tm{x_i}$, and hence no two terms act on dual endpoints. Furthermore, no term $\tm{M_i}$ can perform a communication action by itself, since those are excluded by the definition of actions:
\begin{definition}[Normal Forms]
  A~configuration $\tm{\conf{C}}$ is in normal form if it is of the following form, where each $\tm{M_i}$ is ready to act on $\tm{x_i}$:
  \[
    \tm{\conf{C}}
    \equiv
    \tm{\res{x_1}{x'_1}{\dots\res{x_n}{x'_n}{(\child\;M_1\parallel\dots\parallel\child\;M_m\parallel\main\;V)}}},
  \]
\end{definition}
\begin{restatabletheorem}{thmpgvclosedprogressconfs}[Closed Progress, $\cred$]
  \label{lem:pgv-closed-progress-confs}
  If $\cseq[\main]{\emptyenv}{\conf{C}}$ and $\tm{\conf{C}}$ is in canonical form, then either $\tm{\conf{C}}\cred\tm{\conf{D}}$ for some $\tm{\conf{D}}$; or $\tm{\conf{C}}$ is equivalent to a configuration in normal form.
\end{restatabletheorem}
\input{prf/thm-pgv-closed-progress-confs}
